---
title: "Clustering"
author: "Rishi Shanthan Bhagavatham"
date: "2024-09-10"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# This step is to load the USArrests data
data("USArrests")
head(USArrests)

#The Mean of each Feature
apply(USArrests , 2 , mean)
#The Variance of each Feature
apply(USArrests , 2 , var)
```

a)
```{r}
#First we need to compute the Distance Matrix
dist_matrix <- dist(USArrests, method = "euclidean")

#Perform hierarchical clustering using complete linkage
hc_complete <- hclust(dist_matrix, method = "complete")

#Then Plot the dendrogram
plot(hc_complete, main = "Hierarchical Clustering Dendrogram (Complete Linkage)", xlab = "", sub = "", labels = rownames(USArrests), cex = 0.7)

```


b)
```{r}
#First cut the dendrogram to form three clusters
clusters <- cutree(hc_complete, k = 3)

#Create a data frame with the cluster assignments
clustered_states <- data.frame(State = rownames(USArrests), Cluster = clusters)
clustered_states <- clustered_states[order(clustered_states$Cluster), ]


#Display the states in each cluster
cluster_list <- split(clustered_states$State, clustered_states$Cluster)
cluster_list
```

c)
```{r}

#Scale the variables to have standard deviation one
scaled_data <- scale(USArrests)

#Then we need to compute the Distance Matrix
dist_matrix_scaled <- dist(scaled_data, method = "euclidean")

#Perform hierarchical clustering using complete linkage
hc_complete_scaled <- hclust(dist_matrix_scaled, method = "complete")

#Then Plot the dendrogram
plot(hc_complete_scaled, main = "Hierarchical Clustering Dendrogram (Complete Linkage, Scaled Data)", xlab = "", sub = "", labels = rownames(USArrests), cex = 0.7)

```


d)
```{r}
#Lets compare the dentrograms of a three cluster group from the scaled data for a reference
#So that cut the dendrogram of the scaled data to form three clusters
clusters_scaled <- cutree(hc_complete_scaled, k = 3)

#Create a data frame with the cluster assignments
clustered_states_scaled <- data.frame(State = rownames(USArrests), Scaled_Cluster = clusters_scaled)
clustered_states_scaled <- clustered_states_scaled[order(clustered_states_scaled$Scaled_Cluster), ]

# Display the states in each cluster for scaled data
scaled_cluster_list <- split(clustered_states_scaled$State, clustered_states_scaled$Scaled_Cluster)
scaled_cluster_list

#Compare the original and scaled clustering results
comparison <- merge(clustered_states, clustered_states_scaled, by = "State")
comparison <- comparison[order(comparison$Cluster), ]
comparison
```
It's clear that the clusters are different among the scaled and the regular data. Generally, scaling the variables ensures that each variable contributes equally to the distance calculations. Without scaling, variables with larger variances can dominate the clustering process, potentially leading to misleading clusters. In the USArrests dataset, the variables have different ranges, so scaling is generally recommended to prevent any single variable from having an outsized influence on the clustering outcome.