---
title: "time-series"
author: "Rishi Shanthan Bhagavatham"
date: "2024-10-08"
output:
  pdf_document: default
  html_document: default
---
# R Commands
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exhibit 1.1 on page 2.
```{r}
library(TSA)
```
```{r}
quartz(width=4.875,height=2.5,pointsize=8)
```
```{r}
data(larain);
plot(larain,ylab='Inches',xlab='Year',type='o')
```

## Exhibit 1.2 on page 2.
```{r}
quartz(width=3,height=3,pointsize=8)
```
```{r}
plot(y=larain,x=zlag(larain),ylab='Inches', xlab='Previous Year Inches')
```

## Exhibit 1.3 on page 3.
```{r}
data(color)
plot(color,ylab='Color Property',xlab='Batch',type='o')
```

```{r}
plot(color, color, ylab='Color Property',
xlab='Batch',type='o')
```

```{r}
as.vector(color)
```

```{r}
plot(as.vector(color), color, ylab='Color Property', xlab='Batch',type='o')
```

## Exhibit 1.4 on page 4.
```{r}
plot(y=color,x=zlag(color),ylab='Color Property', xlab='Previous Batch Color Property')
```

## Exhibit 1.9 on page 7.
```{r}
data(oilfilters)
plot(oilfilters,type='l',ylab='Sales')
```

```{r}
Month=c('J','A','S','O','N','D','J','F','M','A','M','J')
plot(oilfilters,type='l',ylab='Sales')
points(y=oilfilters,x=time(oilfilters),pch=as.vector(season(oilfilters)))
```
```{r}
y=rnorm(48)
```
```{r}
plot(y, type='p', ylab='IID Normal Data')
```

```{r}
plot(y, type='o', ylab='IID Normal Data')
```

## Exhibit 2.1 on page 14.
```{r}
n=60
set.seed(12345)
sim.random.walk=ts(cumsum(rnorm(n)),freq=1,start=1)
plot(sim.random.walk,type='o',ylab='Another Random Walk')
```

## Exhibit 3.1 on page 31.
```{r}
data(rwalk)
model1=lm(rwalk~time(rwalk))
summary(model1)
```

# Exhibit 3.2 on page 31.
```{r}
plot(rwalk,type='o',ylab='y')
abline(model1)
```

```{r}
model1a=lm(rwalk~time(rwalk)+I(time(rwalk)^2))
summary(model1a)
```

## Exhibit 3.3 on page 32.
```{r}
data(tempdub)
month.=season(tempdub)
sex=factor(c('M','F','M','M','F'),levels=c('M','F'))
sex
```
```{r}
table(sex)
```

```{r}
model2=lm(tempdub~month.-1)
summary(model2)
```

```{r}
fitted(model2)
```

```{r}
residuals(model2)
```

## Exhibit 3.4 on page 33.
```{r}
model3=lm(tempdub~month.) # intercept is automatically included so one month (January) is dropped
summary(model3)
```
## Exhibit 3.5 on page 35.
```{r}
har.=harmonic(tempdub,1)
model4=lm(tempdub~har.)
summary(model4)
```

```{r}
M=matrix(1:6,ncol=2)
M
```

```{r}
dim(M)
```

```{r}
apply(M,2,mean)
```

## Exhibit 3.6 on page 35.
```{r}
plot(ts(fitted(model4),freq=12,start=c(1964,1)),
ylab='Temperature',type='l',
ylim=range(c(fitted(model4),tempdub)))
points(tempdub)
```

## Exhibit 3.8 on page 43.
```{r}
plot(y=rstudent(model3),x=as.vector(time(tempdub)), xlab='Time', ylab='Standardized Residuals',type='o')
```

## Exhibit 3.11 on page 45.
```{r}
hist(rstudent(model3),xlab='Standardized Residuals')
```

## Exhibit 3.12 on page 45.
```{r}
qqnorm(rstudent(model3))
```

## Exhibit 3.13 on page 47.
```{r}
acf(rstudent(model3))
```

```{r}
shapiro.test(rstudent(model3))
runs(rstudent(model3))
```

# Exercises 1.3-1.5
## 1.3 Simulate a completely random process of length 48 with independent, normal values. Plot the time series plot. Does it look “random”? Repeat this exercise several times with a new simulation each time.
```{r}
par(mfrow = c(3, 1))  # Plot 3 graphs in 1 column

for (i in 1:4) {
  random_process <- rnorm(48)
  
  plot(random_process, type = "l", main = paste("Simulation", i), ylab = "Values", xlab = "Time")
}

par(mfrow = c(1, 1))
```
As we can clearly see, the process looks completely random, with the data oscillating evenly around zero and lacking any clear pattern. 
```{r}
plot(1:48, type = "n", main = "Multiple Random Simulations", ylab = "Values", xlab = "Time", ylim = c(-3, 3))

colors <- rainbow(3)

for (i in 1:4) {
  random_process <- rnorm(48)
  
  lines(random_process, col = colors[i], lwd = 2)
}

legend("topright", legend = paste("Simulation", 1:4), col = colors, lwd = 2)
```

## 1.4. Simulate a completely random process of length 48 with independent, chisquare distributed values, each with 2 degrees of freedom. Display the time series plot. Does it look “random” and non-normal? Repeat this exercise several times with a new simulation each time.

```{r}
length <- 48
df <- 2

simulate_chisquare_process <- function(length, df) {
  rchisq(length, df)
}

plot_time_series <- function(data, title) {
  plot(data, type = "o", col = "blue", main = title, 
       xlab = "Time", ylab = "Value", pch = 16)
  grid()
}

for (i in 1:4) {
  time_series_data <- simulate_chisquare_process(length, df)
  plot_time_series(time_series_data, paste("Chi-Square Distributed Process (2 DoF) - Simulation", i))
}

```
As we can clearly see, the process looks completely random and non-normal, the plot displays only positive values, frequently biased towards the upper end, suggesting a non-normal distribution. While there is an element of randomness, its appearance is notably affected by the skew.
```{r}
plot(1:length, type = "n", ylim = c(0, 10), xlab = "Time", 
     ylab = "Value", main = "Combined Chi-Square Distributed Processes (2 DoF)")
grid()

colors <- rainbow(4)

for (i in 1:4) {
  time_series_data <- simulate_chisquare_process(length, df)
  lines(1:length, time_series_data, type = "o", col = colors[i], pch = 16, lty = i)
}

legend("topright", legend = paste("Simulation", 1:4), col = colors, lty = 1:4, pch = 16)
```

## 1.5. Simulate a completely random process of length 48 with independent, tdistrib-uted values each with 5 degrees of freedom. Construct the time series plot. Does it look “random” and non-normal? Repeat this exercise several times with a new simulation each time.

```{r}
par(mfrow = c(3, 1))  

for (i in 1:3) {
  random_process <- rt(48, df = 5)
  
  plot(random_process, type = "l", main = paste("t-distribution Simulation", i), ylab = "Values", xlab = "Time")
}

par(mfrow = c(1, 1))
```
As it is evident the values are totally random and non- normal, the plot resembles the normal distribution but exhibits more pronounced fluctuations because of the heavier tails of the t-distribution. It appears random, albeit with larger occasional spikes
```{r}

colors <- rainbow(3)

plot(1:48, type = "n", main = "Combined t-distribution Simulations", ylab = "Values", xlab = "Time", ylim = c(-6, 6))

for (i in 1:3) {
  random_process <- rt(48, df = 5)
  

  lines(random_process, col = colors[i], lwd = 2)
}

legend("topright", legend = paste("Simulation", 1:3), col = colors, lwd = 2)
par(mfrow = c(1, 1))

```