---
title: "Auto College Regression Classification"
author: "Rishi Shanthan Bhagavatham"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ISLR2)
library(ggplot2)
```

In case you need to write math expression, please use the quick tutorial as the reference: https://www1.cmc.edu/pages/faculty/aaksoy/latex/latexthree.html

# Question 1

[5 pts] Describe a situation or problem from your job, everyday life, current events, etc., for which a classification model would be appropriate. List some (up to 5) predictors that you might use.

**Solution to Question 1**

`put your solution here `


# Question 2. 

[5 pts] What's the main difference between supervised and unsupervised learning? Give one benefit and drawback for supervised and unsupervised learning, respectively.

**Solution to Question 2**

`put your solution here `


# Question 3 [18 pts]

This question relates to the College data set, which can be found in the file College.csv in the folder. It contains a number of variables for 777 different universities and colleges in the US. The variables are

-	Private : Public/private indicator
-	Apps : Number of applications received
-	Accept : Number of applicants accepted
-	Enroll : Number of new students enrolled
-	Top10perc : New students from top 10 % of high school class
-	Top25perc : New students from top 25 % of high school class
-	F.Undergrad : Number of full-time undergraduates
-	P.Undergrad : Number of part-time undergraduates
-	Outstate : Out-of-state tuition
-	Room.Board : Room and board costs
-	Books : Estimated book costs
-	Personal : Estimated personal spending
-	PhD : Percent of faculty with Ph.D.’s
-	Terminal : Percent of faculty with terminal degree
-	S.F.Ratio : Student/faculty ratio
-	perc.alumni : Percent of alumni who donate
-	Expend : Instructional expenditure per student
-	Grad.Rate : Graduation rate

Before reading the data into R, it can be viewed in Excel or a text editor.

- (a). [2 pts] Use the read.csv() function to read the data into R. Call the loaded data college. Make sure that you have the directory set to the correct location for the data. 

**Solution to Question 3(a)**
`put your solution here `

```{r, echo=TRUE}
# your code

```

- (b). [2 pts]	Look at the data using the View() function. You should notice that the first column is just the name of each university. We don’t really want R to treat this as data. However, it may be handy to have these names for later. Try the following commands:

- `rownames(college) <- college[,1 ]`

- `View(college)`

You should see that there is now a row.names column with the name of each   
university recorded. This means that R has given each row a name corresponding
to the appropriate university. R will not try to perform calculations on the row 
names. However, we still need to eliminate the first column in the data where the 
names are stored. Try

` >college <- college[, -1]`

` > View(college)`

Now you should see that the first data column is Private. Note that another column labeled row.names now appears before the Private column. However, this is not a data column but rather the name that R is giving to each row.

**Solution to Question 3(b)**

`put your solution here `

```{r}
# your code here
```

- (c).
   
  - (i). [2 pts]  Use the `summary()` function to produce a numerical summary of the variables in the data set.
  
   ```{r}
   # your code here
   ```
   
   - (ii). [2 pts] Use the `pairs()` function to produce a scatterplot matrix of the first ten columns or variables of the data. Recall that you can reference the first ten columns of a matrix A using A[,1:10]. Briefly state your observations.

```{r}
# your code here
```

   - (iii). [3 pts] Use the `plot()` function to produce side-by-side boxplots of Outstate versus Private. Briefly state your observations.

```{r}
# your code here
```

  - (iv).[2 pts] Create a new qualitative variable, called Elite, by binning the Top10perc variable. We are going to divide universities into two groups based on whether or not the proportion of students coming from the top 10 % of their high school
classes exceeds 50 %.

`> Elite <- rep("No", nrow(college))`

`> Elite[college$Top10perc > 50] <- "Yes"`

`> Elite <- as.factor(Elite)`

`> college <- data.frame(college , Elite)`

Use the `summary()` function to see how many elite universities
there are. Now use the `plot()` function to produce
side-by-side boxplots of Outstate versus Elite.

```{r}
# your code here
```

  - (v)[2 pts] Use the hist() function to produce some histograms with differing numbers of bins for a few of the quantitative variables. You may find the command `par(mfrow = c(2, 2))` useful: it will divide the print window into four regions so that four plots can be made simultaneously. Modifying the arguments to this function will divide the screen in other ways.

```{r}
# your code here
```

  - (vi)[3 pts]. Continue exploring the data, and provide a brief summary
of what you discover.



# Question 4 [16 pts]
This exercise involves the Boston housing data set.

   -  (a).[2 pts] To begin, load in the Boston data set. The Boston data set is part of the ISLR2 library.

`> library(ISLR2)`

Now the data set is contained in the object Boston.

`> Boston`

Read about the data set:

`> ?Boston`

How many rows are in this data set? How many columns? What do the rows and columns represent?

```{r}
# your code here
```

- (b).[2 pts] Make some pairwise scatterplots of the predictors (columns) in this data set. Describe your findings.

```{r}
# your code here
```

- (c).[2 pts] Are any of the predictors associated with per capita crime rate? If so, explain the relationship.



- (d)[2 pts] Do any of the census tracts of Boston appear to have particularly high crime rates? Tax rates? Pupil-teacher ratios? Comment on
the range of each predictor.


- (e)[1 pt] How many of the census tracts in this data set bound the Charles river?


- (f)[1 pt] What is the median pupil-teacher ratio among the towns in this data set?


- (g)[3 pts] Which census tract of Boston has lowest median value of owneroccupied homes? What are the values of the other predictors for that census tract, and how do those values compare to the overall ranges for those predictors? Comment on your findings.


- (h)[3 pts] In this data set, how many of the census tracts average more than seven rooms per dwelling? More than eight rooms per dwelling? Comment on the census tracts that average more than eight rooms per dwelling.




# Question 5. [6 pts]
Suppose we have a data set with five predictors, $X1 = \text{GPA}, X2 =IQ, X3 = \text{Level (1 for College and 0 for High School)}, X4 =\text{Interaction between GPA and IQ, and }X5 = \text{Interaction between GPA and Level.}$ The response is starting salary after graduation (in thousands of dollars). Suppose we use least squares to fit the model, and get $\hat{\beta}_0 = 50, \hat{\beta}_1 = 20, \hat{\beta}_2 = 0.07, \hat{\beta}_3 = 35, \hat{\beta}_4 = 0.01, \hat{\beta}_5 = −10$.

- (a)[3 pts] Predict the salary of a college graduate with IQ of $110$ and a GPA of $4.0$.

```{r}
# your code here
```

- (b)[3 pts] True or false: Since the coefficient for the GPA/IQ interaction term is very small, there is very little evidence of an interaction effect. Justify your answer.

```{r}
# your code here
```

# Question 6. [34 pts]
This question involves the use of multiple linear regression on the
Auto data set.
```{r}
data(Auto)
data<-Auto
head(data)
```

- (a).[3 pts] Produce a scatterplot matrix which includes all numeric variables
in the data set. Please give the interpretation of the plot.

```{r}
# your code here

```
- (b) [3 pts] Compute the matrix of correlations between the variables. You will need to exclude the name variable, which is qualitative. Explain your discoveries, find the top 1 predictors.

```{r}
# your code here

```

- (c)[6 pts] perform a simple linear regression use MPG as response and the top predictor. Produce diagnostic plots of the linear regression fit (similar as the ones we showed on class to check the assumptions of linear regression: linearity, normality, constant variance). Interpret your discoveries.

```{r}
# your code here

```




- (d)[4 pts] Use the `lm()` function to perform a multiple linear regression with mpg as the response and all other variables except name as the predictors. Use the summary() function to print the results. Comment on the output. For instance:

  - i. Is there a relationship between the predictors and the response?
  
  - ii. Which predictors appear to have a statistically significant relationship to the response?
  
  - iii. What does the coefficient for the year variable suggest?

```{r}
# your code here

```
  
  
- (e)[5 pts] Produce diagnostic plots of the linear regression fit. Comment on any problems you see with the fit. Do the residual plots suggest any unusually large outliers? Does the leverage plot identify any observations with unusually high leverage?

```{r}
# your code here

```
  
- (f) [5 pts] Fit a multiple linear regression model called model_tops using MPG as the response and the top 3 predicting variables with the strongest relationship with MPG. Compare the coefficients of the 3 predictor in model_tops with the model that used all numeric predictors, which ones are larger?  Why? 
 
```{r}
# your code here

``` 

- (g)[3 pts] Use the * and : symbols to fit linear regression models with interaction effects. Do any interactions appear to be statistically
significant?

```{r}
# your code here

``` 

- (h)[5 pts] Try a few different transformations of the variables, such as $\log(X), \sqrt{X}, X^2$. Comment on your findings.

```{r}
# your code here

``` 